---
title: "Summaries"
author: "Kaisa Roots"
date: "2019-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(nycflights13)
```

dplyr summarise() function can be used to calculate counts and proportions of logical values: sum(x > 10), mean(y == 0). 

When used with numeric functions, TRUE is converted to 1 and FALSE to 0. 

> This makes sum() and mean() very useful: sum(x) gives the number of TRUEs in x, and mean(x) gives the proportion.

### Exercises

1. Using nycflights13 "flights" dataset, brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. 

```{r}
flights
```

Consider the following scenarios:

  - A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.

```{r}
flights %>% 
  group_by(flight) %>% 
  summarize(n = n(),
            fifteen_early = mean(arr_delay == -15, na.rm = TRUE),
            fifteen_late = mean(arr_delay == 15, na.rm =TRUE)) %>% 
  filter(fifteen_early == 0.5 | fifteen_late == 0.5)
```

  - A flight is always 10 minutes late.

```{r}
flights %>% 
  group_by(flight) %>% 
  summarize(n = n(),
            ten_late = mean(arr_delay == 10, na.rm = TRUE)) %>%
  filter(ten_late == 1)
```

  - A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.
  
```{r}
flights %>% 
  group_by(flight) %>% 
  summarize(n = n(),
            thirty_early = mean(arr_delay == -30, na.rm = TRUE),
            thirty_late = mean(arr_delay == 30, na.rm =TRUE)) %>% 
  filter(thirty_early == 0.5 & thirty_late == 0.5)
```

  - 99% of the time a flight is on time. 1% of the time it’s 2 hours late.

```{r}
flights %>% 
  group_by(flight) %>% 
  summarize(n = n(),
            percentage_on_time = mean(arr_delay == 0, na.rm = TRUE),
            two_hours = mean(arr_delay > 120, na.rm = TRUE)) %>% 
  filter(percentage_on_time == 0.99 & two_hours == 0.01)
```

  - Which is more important: arrival delay or departure delay?
  
  Both arrival delay and departure delay are unwanted. With arrival delay, many other flights are affected and connected flights might be missed. However, departure delay directly contibutes to arrival delay. Additionally, the difference between arrival delay and departure delay are used to define air time, hence real arrival time can show whether any problems during the flight have occured. 

2. Come up with another approach that will give you the same output as not_cancelled %>% count(dest) and not_cancelled %>% count(tailnum, wt = distance) (without using count()).

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
not_cancelled
```

##### 1

```{r}
not_cancelled %>% 
  count(dest)
```

```{r}
not_cancelled %>%
  group_by(dest) %>% 
  summarize(n = n())
```

##### 2

```{r}
not_cancelled %>% 
  count(tailnum, wt = distance)
```

```{r}
not_cancelled %>% 
  group_by(tailnum) %>%
  summarize(n = sum(distance))
```

3. Our definition of cancelled flights (is.na(dep_delay) | is.na(arr_delay) ) is slightly suboptimal. Why? Which is the most important column?

We assume that if a flight did not leave it was cancelled. Therefore, when the condition is.na(dep_delay) is met, the flight was cancelled. 

4. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

```{r}
flights %>% 
  group_by(day) %>% 
  summarize(cancelled = mean(is.na(dep_delay)))
```

```{r}
flights %>% 
  group_by(day) %>% 
  summarize(cancelled = mean(is.na(dep_delay)),
            dep_del = mean(dep_delay, na.rm = TRUE),
            arr_del = mean(arr_delay, na.rm = TRUE)) %>% 
  ggplot(aes(y = cancelled)) +
  geom_point(aes(x = dep_del), colour = "pink") +
  geom_point(aes(x = arr_del), colour = "green") +
  labs(x = "Average delay per day", y = "cancelled flights per day")
```

This plot shows us that there are more cancelled flights on days with more departure and arrival delays, which is logical, because a delay of a flight affects other flights as well. We also see that there are more departure delays (pink) on an average day. 

5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))

```{r}
flights %>% 
  group_by(carrier) %>% 
  summarize(n = n(),
            dep_max = max(dep_delay, na.rm=TRUE),
            arr_max = max(arr_delay, na.rm=TRUE)) %>% 
  arrange(desc(dep_max, arr_max)) %>% 
  filter(1 : n() == 1)
```

```{r}
flights %>% 
  group_by(carrier, dest) %>% 
  summarize(n = n(),
            dep_max = max(dep_delay, na.rm=TRUE),
            arr_max = max(arr_delay, na.rm=TRUE)) %>% 
  arrange(desc(dep_max, arr_max)) %>% 
  filter(1 : n() == 1)
```

6. What does the sort argument to count() do. When might you use it?

If TRUE, 'sort' will sort count() output in descending order of n. 

```{r}
flights %>% 
  count(carrier, sort = TRUE)
```

```{r}
flights %>% 
  count(flight, sort = TRUE)
```
